---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# MetaboMate

The goal of MetaboMate is to ...

## Installation

The MetaboMate package is currently hosted on GitHub. Installing it requires the developer tools R package:
```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("kimsche/MetaboMate")
```
The installation process can take up to about 3 minutes as the tutorials are build locally with high-resolution NMR example data.

&nbsp;


#### Installation Notes for R Starters

There might come up **warning messages** (eg. *unknown time zone*). These are not a problem! :pray:
However, **error messages** are a problem as these stop the installation process. 

&nbsp;

#### Solutions to Common Installation Errors:

>`Installation failed: 'exdir' does not exist`

Workspace directory not writable, change the location of the current workspace

>`Dependency package missing`

Install the required package(s) from Bioconductor.org

&nbsp;

You can check if the installation was successful by loading the package into your R session: `library(MetaboMate)` - all is fine if no error message is returned.





## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```
