---
title: "NMR-based Metabolic Profiling with R - Data Import and Preprocessing"
author: "Torben Kimhofer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette documents a typical pre-processing workflow for NMR-based metabolic profilig using the *MetaboMate* R package. Example data represent Proton NMR spectra of [how many] murine urine samples collected pre and post bariatric surgery. One dimensional spectra were acquired on a 600 MHz Bruker Avance III spectrometer, equipped with a 5 mm triple resonance (TXI) probe operating at 300 K. Further information on sample collection and processing as well as data acquisition can be obtained from the original publication by Jia V. Li *et al.*^[Li, Jia V., *et al.* (2011) Metaboilc surgery profoundly influences gut microbial-host mtaboilc cross-talk. *Gut* 60.9, 1214-1223.]

## Prerequisites
Although not essential for this tutorial, I recommend to install RStudio. RStudio is an open-source integrated development environment for R, which includes a code editor that highlights syntax, enables quick and easy access to help pages, improves workspace management and offers various tools for plotting, history and debugging. Simply put, it makes working with R a lot more efficient.

## Reading 1D NMR spectra 
The first step is to load NMR spectra into the R workspace. This can be accomplished with the **readBruker()** function, which imput argument is the path to the NMR experiment folder. It then imports all spectra into matrix format where each row represents an observation (here this is a urine NMR spectrum) and each column represents a metabolic feature (ppm variable).
```{r, fig.show='hold'}
library(MetaboMate)
#path=system.file("extdata/", package = "MetaboMate")
path='../inst/extdata/'
readBruker(path)
ls()
```

The function **readBruker()** searches for all experiment files in a given path and automatically adds the NMR data matrix, ppm and spectrometer metadata (as respective variables X, ppm  and meta) to the R workspace. The dataframe **meta** contains detailed information about acquisition and processing parameters for each spectrum (rows=specta, columns=parameters), as extracted from the Bruker *acqus* and *procs* files. The prefix a or p indicates parameter was extracted from the acqus or procs file, respectively. For example, the column **a_RG** contains the receiver gain vale for each spectrum, which was extracted from the experiment's **acqus** file.

## Plotting spectra 
For the visualisation of NMR spectra there are two low-level plotting functions (**spec()** and **matspec()**), which allow plotting an individual spectrum or multiple spectra, respectively. Both of these functions are fairly fast. Let's have a first look at all imported spectra over the entire ppm range with the **matspec()** function.

```{r, fig.show='hold', fig.width = 7.2, fig.height = 4}
matspec(ppm, X, shift = range(ppm))
```

From this overview we can see that the spectral width ranges from -5 to approximately 15 ppm with the residual water signal resonating around 4.8 ppm. 

There are different higher-level plotting functions available that allow a more comprehensive spectral visualisation of individual peaks or peak areas. One of this function is **specOverlay**, which is comparable to **matspec()** shown above. For illustration purposes we plot the TSP signal (a component of the sample preparation buffer) coloured accroding to the receiver gain paramter in **meta** dataframe. The RG value amplifies the NMR signal (free induction decay) before it is converted to a digital signal. 

```{r, fig.show='hold', fig.width = 7.2, fig.height = 4}
# plot TSP signal
specOverlay(X, ppm, shift=c(-0.05,0.05), 
            an=list('All_Spectra'='All spectra', # facetting (here: a single panel) 
                    'Receiver Gain'=as.numeric(meta$a_RG), # colour
                    'Experiment'=factor(meta$a_EXP))) # linetype

```

From the plot above you can see that there were different types of experiments performed (indicated by different linetypes) and that the RG is related to the signal strength in each experiment. 


The different experiment types map to different biolfuids and NMR pulse sequences, which makes quantitative comparison across spectra not possible. Therefore, we must filter for the desired experiment type (**<JL-noe1d>**) before we continue with the analysis.

```{r, fig.show='hold', fig.width = 7.2, fig.height = 4}
# plot TSP signal
specOverlay(X, ppm, shift=c(-0.05,0.05), 
            an=list('All_Spectra'='All spectra', # facetting (here: a single panel) 
                    'Receiver Gain'=as.numeric(meta$a_RG), # colour
                    'Experiment'=factor(meta$a_EXP))) # linetype

```


## Phasing and calibration
Phasing script not available but work in progress.

Spectral calibration is an essential step in NMR data processing, where the entire spectrum is shifted until the peak apes of a reference compound is at a certain ppm position. For urine NMR analysis the reference compounds is usually TSP, which is calibrated at zero ppm.^[Dona, A.C., *et al.* (2014) Precision high-throughput proton NMR spectroscopy of human urine, serum, and plasma for large-scale metabolic phenotyping. *Analytical Chemistry*. 86.19. 9887-94.]

We can calibrate the urine spectra with the following code:
```{r, fig.show='hold', fig.width = 7.2, fig.height = 4}
# calibrate urinary NMR spectra to TSP
X.cali=calibration(X, ppm, type='Urine')

# plot TSP overlay with calibrated data
specOverlay(X.cali, ppm, shift=c(-0.05,0.05), 
            an=list('All_Spectra'=1 , 
                    'Receiver Gain'=as.numeric(meta$a_RG),
                    'Experiment'=factor(meta$a_EXP)))

ls()
```

Now you can see that all TSP signals are nicely aligned with the peak apex centered at zero ppm. By removing non-relevant experiments we can clearly see that the receiver gain (RG) parameter has pronounced effect on specral intensities. In metabolic phenotyping, this parameter is often fixed for spectra acquited within one analytical run as it sometimes can be diffucult to account for. 


## Assessment of spectral quality
In metabolic phenotyping and in any other field where multiple NMR spectra are compared quantitatively, the quality of spectra is of particular importance. Assessment of spectral quality usually is a visual inspection of the water suppresion quality, spectral line widths and baseline stability as well as the average signal to noise (SN) ratio.

High throughput NMR often does not allow a manual inspection of each individual spectrum and automatically generated quality indices are often used to exclude low quality spectra. Here, we derive several spectral quality indices:

```{r, fig.show='hold', fig.width = 7.2, fig.height = 4}
# calculate quality control measures
spec.qc=spec.quality(X.cali, ppm, ppm.noise=c(9.4,9.5), plot=T)
knitr::kable(head(spec.qc, 6))
```

In the above plot we see that ...

Currently, the quality control line width estimate requires a TSP signal or any other external reference compound that resonates at zero ppm.

## Excision chemical shifts regions and baseline correction
 
```{r, fig.show='hold', fig.width = 7.2, fig.height = 4}
# Indexing regions of TSP signal ...
idx.TSP=get.idx(c(min(ppm), 0.5), ppm)
# ...residual water signal
idx.water=get.idx(c(4.6, 5), ppm)
# ... as well as ppm regions without any signals
idx.noiseDF=get.idx(c(9.5, max(ppm)), ppm)
# Excising these regions
X.cali=X.cali[,-c(idx.TSP, idx.water, idx.noiseDF)]
ppm=ppm[-c(idx.TSP, idx.water, idx.noiseDF)]

# Baseline correction
X.bl=bline(X.cali)

# compare spectra before and after baseline correction
matspec(ppm, X.cali, shift=c(3.6,4), main='Raw')
matspec(ppm, X.bl, shift=c(3.6,4), main='Baseline corrected')
```

## Spectral normalisation
Depending on the sample type, multivariate analysis requires normalisation of spectra. In our urinary NMR study for exmample, the urine sample dilution varies across samples, perhaps due to the uptake of different amounts of water.

There are several different normalisation methods avaiable. Among the most commonly applied are total area normalisaton and probablistic quotient normalisation.^[Dieterly, F., \emph{et al.} (2006), Probabilistic Quotient Normalization as Robust Method to Account for Dilution of Complex Biological Mixtures. Application in 1H NMR Metabonomics, \emph{Analytical Chemistry}, 78.3, 4281-90] In this tutorial, we normalise the samples using both methods for comparative purposes.

```{r, fig.show='hold'}
X.pqn=pqn(X.bl, add.DilF = 'dilF.pqn', TArea = F)
X.pqn.ta=pqn(X.bl, add.DilF = 'dilF.pqn1', TArea = T)
X.ta=totalArea(X.bl, add.DilF='dilF.ta')

plot(dilF.pqn, dilF.pqn1)
# plot(dilF.pqn1, dilF.ta)
plot(dilF.pqn, dilF.ta)
abline(a = c(0,1), lty=2)
```

In the present case PQN generates on average slightly lower dilution factors than total area normalisation (the dotted line indicates perfect correlation) with increasing differences towards higher dilutions. Normalisation scales an entire spectrum up or down (multiplication with the dilution factor), and improper dilution factor normalisation can have detrimental effects for further downstream analyses. 

